Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-08-18 paper/letter.tex
latest version of the letter. no black hole.

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Lyman-$\alpha$ sketches a massive dwarf}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of $71^{+12}_{-12}$ km s$^{-1}$ with outflows of
$55^{+67}_{-37}$ km s$^{-1}$) consistent with a dynamical mass on
the order of two billion solar masses, which is at least $20$ times
larger than the neutral mass hydrogen in the galaxy.   
We argue that a plaussible explanation for the high dynamical mass is
the presence of a super-massive black hole.  
This work demonstrates the importance of considering rotation and
multiphase physics among the possible conditions shaping the
\lya\ spectra of the first galaxies. 
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
}

@@

The best parameters in the rotation model are a rotational velocity of 
$V_{\rm max}=348^{+75}_{-48}$ \kms, an optical depth $\log_{10}\tau=6.96^{+0.26}_{-0.18}$,  and a temperature of $\log_{10} T/\mathrm {K} = 4.27^{+0.11}_{-0.18}$. 
This translates into a column density of $\log N_{HI} /
\mathrm{atoms\ cm}^{-2} =  20.5^{+xx}_{-xx}$.  
This model is also able to constrain the angle between the plane perpendicular to the rotation axis and the observational line-of-sight to
$\theta = 35.78^{+2.13}_{-1.88}$ degrees.

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=71^{+12}_{-12}$ \kms,
the clumps outflowing velocity $v_{\infty, \rm{cl}}=55^{+67}_{-37}$
\kms and the fraction of the \lya emission that is  coming
from the cold clumps  
$P_{cl}=0.77^{+0.15}_{-0.44}$. 
The multiphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $M_{\rm
  dyn}=2.9^{+1.1}_{-0.9}\times 10^{9}$ $M_{\odot}$.

In the rotation model assuming reasonable bounds for the number
density of neutral Hydrogen atoms (
$0.01<n/\mathrm{atoms\ cm^{-3}}<0.1$) and using $\tau=10^7$ the radius
of the emission region can be bracketed to be in $0.55 <
r_s/\mathrm{kpc}< 5.5$.
Using $V_{\rm max}=300$km s$^{-1}$, we find a dynamical mass range of
$5.45\times 10^{9}M_{\odot} < M_{\rm dyn} < 5.45\times 10^{10} M_{\odot}$.

Radio surveys of the 21 cm line have put an upper limit to the neutral
hydrogen mass in \tol\ of $M<2.65\times 10^{8}$ M$_{\odot}$
\cite{pustilnikmartin07}.  
Our models thus provide evidence for a dynamical mass at least 20 times
higher than the mass in neutral hydrogen.   
There are three main possibilities to explain the dynamical 
mass: stars, dark matter and a supermassive black hole. 
The stellar mass in old stars has been constrained by non-detections
in the K-band to be $<XXX$M$_{\odot}$.
Considering a standard dark matter halo, it can only contribute to
$xx$ M$_{\odot}$ in the region of interest. 
A super massive black hole remains an open possibility. 
Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total object mass, suggesting that
there is a large population of undetected black holes in dwarf
galaxies.  
Ionization emission lines present in this galaxy support the idea that
\tol\ might have harbored an an Active Galactic Nucleus.

@@

These findings demonstrate the importance of including rotation and multiphase
conditions as features to model the \lya\ line in high redshift
galaxies.
Additionally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistent with future observational
kinematic maps, it could correspond to a so far undetected black hole
in a dwarf galaxy, providing a new way to test and probe
theories on the co-evolution of galaxies and black holes in the first
generation of galaxies.

@@

FIXME: Add non detection in 2MASS %\url{http://vizier.u-strasbg.fr/cgi-bin/VizieR?-source=B/2mass}

\tol receding velocity is $7785\pm 50$km s$^{-1}$, which translates
into a distance of $106.6$ Mpc (Hubble constant 73 Mpc km$^{-1}$
s$^{1}$)
Its metallicity is $\sim Z_{\odot}/24$ \cite{Izotov04} as derived from optical
spectroscopy.

The observed flux for the Lyman alpha line is $\sim
8.1\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \cite{Thuan97}
and a Equivalent Width of $70$\AA and its H$\beta$ flux is 
$1.62\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \AA${-1}$
\cite{Izotov04} which gives a Ly$\alpha$/H$\beta$ flux ratio of
4.9$\pm$0.1. The Ly-$\alpha$ flux values correspond to luminosities of
$L_{Ly\alpha}=2.2\times 10^{42}$ erg s$^{-1}$ over a $20$\AA
bandwidth, which in turns translates  into a star formation rate of
$2.0$ M$_{\odot}$ yr$^{-1}$ using a standard conversion factor between
luminosity and star formation rate of $9.1\times 10^{-43}$
$L_{Ly\alpha}$ M$_{\odot}$ yr$^{-1}$. 
The absolute magnitude in the $V$ band translates into a luminosity of
$8.9\times 10^{8}$ L$_{\odot}$.
% using http://tomdwelly.com/tools_fluxtolum.php
Comparing this ratio with the theoretical expectation from case B
recombination of $23.3$ \cite{Hummer1987} one can estimate an escape
fraction of $20$\% for Ly$\alpha$ radiation.

@@

Interpretation by \cite{mashesse03}.

@@

From the multiphase model we obtained $v=71_{-25}^{+17}$ \kms
and $r=5$ kpc. 
This corresponds to a dynamical mass of $M_{\rm
  dyn}=5.9^{+3.1}_{-3.4}\times 10^{9}$ $M_{\odot}$ which is at least
$20$ times the HI mass estimated from observations.

>>>>

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Lyman-$\alpha$ emission reveals a dispersion-supported dwarf}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of $69^{+17}_{-15}$ km s$^{-1}$ with outflows of
$79^{+73}_{-42}$ km s$^{-1}$) consistent with a dynamical mass at
least $10$ times larger than its baryonic mass.  
We argue that the most plaussible explanation is provided by the
multiphase model as it is a close match to the galaxy scaling relationship
for a dark matter dominated dwarf elliptical supported by its
velocity dispersion. 
This work demonstrates the importance of considering multiphase
physics among the possible conditions shaping the \lya\ spectra of the
first galaxies.   Additionally, if future kinematic maps of
\tol\ confirm the high velocities postulated in our model, it would
help to stablish the \lya\ line morphology as an observational tracer to
study the co-evolution of the first generation of galaxies and its
host dark matter halos. 
}

@@

The best parameters in the rotation model are a rotational velocity of 
$V_{\rm max}=348^{+75}_{-48}$ \kms, an optical depth
$\log_{10}\tau=6.96^{+0.26}_{-0.18}$,  and a temperature of $\log_{10}
T/\mathrm {K} = 4.27^{+0.11}_{-0.18}$.  
This model is also able to constrain the angle between the plane
perpendicular to the rotation axis and the observational line-of-sight
to $\theta = 35.78^{+2.13}_{-1.88}$ degrees.

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=69^{+17}_{-15}$ \kms,
the clumps outflowing velocity $v_{\infty, \rm{cl}}=79^{+73}_{-42}$
\kms and the fraction of the \lya emission that is  coming
from the cold clumps  $P_{cl}=0.78^{+0.16}_{-0.43}$. 
The multiphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $M_{\rm
  dyn}=2.8^{+1.3}_{-1.2}\times 10^{9}$ $M_{\odot}$. In the rotation
model assuming reasonable bounds for the number density of neutral Hydrogen atoms (
$0.01<n/\mathrm{atoms\ cm^{-3}}<0.1$) and using $\tau=10^7$ the radius
of the emission region can be bracketed to be in $0.55 <
r_s/\mathrm{kpc}< 5.5$.
Using $V_{\rm max}=300$km s$^{-1}$, we find a dynamical mass range of
$5.45\times 10^{9}M_{\odot} < M_{\rm dyn} < 5.45\times 10^{10} M_{\odot}$.

Which solution is more plausible? If \tol\ followed the fundamental
plane relationship between its mean surface brightness $I_e$, the
projected half-light radius $R_e$ and the velocity dispersion
$\sigma$, described by $\log I_e=1.6 \log\sigma - 1.21\log R_e +
0.55$ \cite{2009ApJ...698.1590G,2011ApJ...726..108T},  
the expected velocity dispersion would be on the order of $47\pm 3$
\kms, which only presents a small tension with the multiphase model.   
The solution of a high rotational velocity pushes \tol\ outside the
scaling relationship for dwarfs. One could account for the high
velocities if the galaxy had a supermassive black hole accounting for
the mass that cannot be explained by baryonic or dark mass. The economy of
assumptions makes it more reasonable to expect that the multiphase
model could be a better match to reality.

@@

These findings demonstrate the importance of including multiphase
conditions as a feature to model the \lya\ line in high redshift
galaxies.
If the estimated velocity dispersion proves to be consistent with
future observational kinematic maps, it could provide a new way to
study velocity dispersions in distant gas rich starbursting galaxies, 
opening the possibility to study the dynamical co-evolution of
galaxies and dark matter halos in the first generation of galaxies.

@@

%\url{http://vizier.u-strasbg.fr/cgi-bin/VizieR?-source=B/2mass}
% TOLOLO 00021
%https://ned.ipac.caltech.edu/cgi-bin/objsearch?objname=Tol+21&extend=no&hconst=73&omegam=0.27&omegav=0.73&corr_z=1&out_csys=Equatorial&out_equinox=J2000.0&obj_sort=RA+or+Longitude&of=pre_text&zv_breaker=30000.0&list_limit=5&img_stamp=YES 
\tol receding velocity is $7785\pm 50$km s$^{-1}$, which translates
into a distance of $106.6$ Mpc (Hubble constant 73 Mpc km$^{-1}$
s$^{1}$)
Its metallicity is $\sim Z_{\odot}/24$ \cite{Izotov04} as derived from optical
spectroscopy.

The observed flux for the Lyman alpha line is $\sim
8.1\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \cite{Thuan97}
and a Equivalent Width of $70$\AA and its H$\beta$ flux is 
$1.62\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \AA${-1}$
\cite{Izotov04} which gives a Ly$\alpha$/H$\beta$ flux ratio of
4.9$\pm$0.1. The Ly-$\alpha$ flux values correspond to luminosities of
$L_{Ly\alpha}=2.2\times 10^{42}$ erg s$^{-1}$ over a $20$\AA
bandwidth, which in turns translates  into a star formation rate of
$2.0$ M$_{\odot}$ yr$^{-1}$ using a standard conversion factor between
luminosity and star formation rate of $9.1\times 10^{-43}$
$L_{Ly\alpha}$ M$_{\odot}$ yr$^{-1}$. 
The absolute magnitude in the $V$ band translates into a luminosity of
$8.9\times 10^{8}$ L$_{\odot}$.
% using http://tomdwelly.com/tools_fluxtolum.php
Comparing this ratio with the theoretical expectation from case B
recombination of $23.3$ \cite{Hummer1987} one can estimate an escape
fraction of $20$\% for Ly$\alpha$ radiation.
The bolometric UV luminosity is $9.43\pm1.94 \times 10^{8}$
L$_{\odot}$ as measured by GALEX.

@@

The near-infrared fluxes at $3.6$ $\mu$m and $4.5$ $\mu$m are
$7.71\pm0.55\times 10^{-5}$ Jy and $7.98\pm0.71\times 10^{-5}$ Jy
\cite{2008ApJ...678..804E}.  Using a convertion betwen fluxes and
stellar mass calibrted on the Large Magellanic Cloud $M_{\star} =
10^{5.65} \times F_{3.6}^{2.85} \times F_{4.5}^{-1.85} \times
(D/0.05)^2 M_{\odot}$, where fluxes are in Jy and $D$ is the luminosity
distance to the source in Mpc, we find $M_{\star} = 1.45\pm0.45\times 10^{8}
M_{\odot}$, with a $30\%$ uncertainty coming from the callibration
process \cite{2012AJ....143..139E}.

@@

From the multiphase model we obtained $v=69_{-15}^{+17}$ \kms
and $r=5$ kpc. 
This corresponds to a dynamical mass of $M_{\rm
  dyn}=2.8^{+1.3}_{-1.2}\times 10^{9}$ $M_{\odot}$ which is at least
$5$ times the HI mass plus the stellar mass estimated from
observations.
